### 2024-02-06 README.Rmd
shorten readme

Below, we look at these functions in more detail and show them in operation.

## Pre-functions

`sfdep` offers excellent tools for building neighbourhood structures, among other things. It has a range of functions depending on how we want to define what *neighbour* should mean.

However, often when preparing areal spatial data, the presence of uncontiguous areas (such as islands or exclaves) can create difficulties. We might also want to account for some hidden contiguities by allowing bridges, tunnels etc. to render two uncontiguous areas as neighbours.

`sfislands` provides functions to make this task easier.

It also provides a number of further helper functions to examine these neighbourhood structures and use them in models such that the workflow is streamlined from the human side.

### Mainland France (`guerry` dataset)

`sfislands` is intended to be used alongside the `sfdep` package. The principal functions from `sfdep` for areal data are demonstrated in its vignette using the `guerry` ("*Essay on the Moral Statistics of France*") dataset. This data applies to the geographical area of mainland France as shown below:

```{r}
g <- guerry |> 
  st_as_sf()

ggplot(g) + 
  geom_sf() + 
  theme_void()
```

In this French example, the `st_islands` function **st_bridges()** produces exactly the same neighbourhood structure (based on contiguity) as **sfdep::st_contiguity()**. This can be seen below by piping these neighbourhood outputs through the `st_islands` function **st_quickmap_nb()** which gives a visual representation of the structure on a map. This function works with neighbourhoods constructed from any package as long as they are in list or matrix form.

For **st_quickmap_nb()**, the neighbourhoods should be within an `sf` dataframe as a column called "nb". **st_bridges()** does this automatically whereas the "nb" column needs to be added when using `sfdep` functions.

#### sfislands::st_bridges()

The `st_bridges()` contiguities below, where each department is considered a neighbour of another if it touches it at least one point...

```{r}
g |> st_bridges("department") |> 
  st_quickmap_nb()

... are the same as these from `sfdep`:

#### sfdep::st_contiguity()

```{r}
g |> mutate(nb = st_contiguity(geometry)) |> 
  st_quickmap_nb()

As these maps are produced using `ggplot2`, their characteristics can be edited using normal `ggplot2` syntax (see 1, below). For convenience, simple arguments are also provided (see 2, below) for changing the core characteristics of the map in a simple way:

```{r, fig.width=8}
ggarrange(

  g |> mutate(nb = st_contiguity(geometry)) |> 
    st_quickmap_nb() +
    geom_sf(data = g |> group_by(region) |> summarise(), 
            linewidth = 0.5, colour = "black", fill = NA) + 
    labs(title = "st_quickmap_nb()",
         subtitle = "1. using ggplot syntax") + 
    theme_minimal() +
    theme(panel.background = element_rect(fill = "aquamarine3", color = "black"),
          axis.text = element_blank()),

  g |> mutate(nb = st_contiguity(geometry)) |> 
    st_quickmap_nb(linkcol = "orange", 
                        bordercol = "white", 
                        pointcol = "yellow", 
                        fillcol = "black", 
                        linksize = 0.4, 
                        bordersize = 0.3, 
                        pointsize = 0.8,
                        title = "st_quickmap_nb()",
                        subtitle = "2. using simplified arguments"),

  ncol=2
)

`sfdep` offers a number of different types of neighbourhood structure, a selection of which are shown below. These can again be conveniently visualised using the **st_quickmap_nb()** function:

#### sfdep::st_dist_band()

All areas within a certain distance are considered neighbours:

```{r}

g |> mutate(nb = st_geometry(g) |> 
              st_dist_band(upper = 150000)) |> 
  st_quickmap_nb()

#### sfdep::st_knn()

The k-nearest-neighbours (here, 1) to each area are considered neighbours:

```{r}

g |> mutate(nb = st_geometry(g) |> 
              st_knn(1, symmetric = TRUE)) |> 
  st_quickmap_nb()

#### sfdep::st_block_nb()

All areas within a chosen block are considered neighbours:

```{r}

id <- g$code_dept
regime <- g$region
g |> 
  mutate(
    nb = st_block_nb(regime, id)
  ) |> 
  st_quickmap_nb()

#### sfdep::st_lag_cumul()

Cumulative higher orders of contiguity such as also including neighbours-of-neighbours:

```{r}

g |> mutate(nb = st_contiguity(geometry) |> st_nb_lag_cumul(2)) |> 
  st_quickmap_nb()

However, the `sfdep` functions above which rely on contiguity will run into difficulties if we consider the following geography:

### England, Scotland & Wales

In the context of the constituencies of England, Scotland and Wales, there will be problems due to the presence of islands which will not feature in a contiguity graph unless we attempt to pick out these islands and set up a buffer around them. This process is, however, cumbersome and can induce contiguities which are not intended.

One potential remedy is to entirely exclude islands from the study, another is to construct a contiguity structure according to your desired criteria and then set the islands to be contiguous to their closest *k* constituencies.

`st_bridges()` can do both of these things.

Below is the map in question:

```{r, echo=FALSE}
uk_election$country <- case_when(uk_election$region == "Wales" ~ "Wales",
                                  uk_election$region == "Scotland" ~ "Scotland",
                                  TRUE ~ "England")
```

There are island constituencies in the north around Scotland but also less obvious ones in Wales and England. The constituencies which are non-contiguous are outlined in red below:

```{r, fig.width=12, fig.height=10, echo=FALSE}
ggplot() + 
  geom_sf(data=uk_election, aes(fill=country), colour="white") + 
  # geom_sf_label(data=uk_election |> group_by(country) |> summarise(), aes(label=country),size=10) +
  geom_sf(data=uk_election |> filter(constituency_name == "Na h-Eileanan An Iar"), colour="#B20016", fill=NA, linewidth=0.7) + 
  geom_sf_label(data=uk_election |> filter(constituency_name == "Na h-Eileanan An Iar"),
                aes(label=constituency_name)) +
  geom_sf(data=uk_election |> filter(constituency_name == "Orkney and Shetland"), colour="#B20016", fill=NA, linewidth=0.7) + 
  geom_sf_label(data=uk_election |> filter(constituency_name == "Orkney and Shetland"),
                aes(label=constituency_name), nudge_x = -80000, nudge_y = -60000) +
    geom_sf(data=uk_election |> filter(constituency_name=="Ynys Mon"),colour="#B20016", fill=NA, linewidth=0.7) + 
    geom_sf_label(data=uk_election |> filter(constituency_name=="Ynys Mon"),
                  aes(label=constituency_name), nudge_x = -90000) +
    geom_sf(data=uk_election |> filter(constituency_name=="Isle Of Wight"),colour="#B20016", fill=NA, linewidth=0.7) + 
    geom_sf_label(data=uk_election |> filter(constituency_name=="Isle Of Wight"),
                  aes(label=constituency_name), nudge_x = 50000, nudge_y = -30000) +
  scale_fill_manual(values = c("antiquewhite4","#1C4769", "#24975E")) +
  # guides(fill = "none") + 
  theme_void()
```

To incorporate these we use `st_bridges()`. We can set `remove_islands` to TRUE if we decide to simply exclude the islands, or we set `link_k_islands` to the closest k constituencies to each islands which we want to bridge.

#### Remove islands

Below, with the argument `remove_islands` set to TRUE, we simply remove these islands from the dataset entirely.

```{r, fig.width=12, fig.height=10}
nbsf <- st_bridges(df = uk_election,
                   geom_col_name = "constituency_name",
                   remove_islands = T)
st_quickmap_nb(nbsf,
                    pointsize=0.05,
                    title = "st_bridges() contiguities",
                    subtitle = "no island constituencies\n(islands which remain are part of a contiguous constituency)")

#### Connect islands

Alternatively, we can join islands to the nearest, say, 2 constituencies. `st_bridges()` by default returns the original dataframe augmented with a "nb" column which contains the contiguities in list form.

```{r, fig.width=12, fig.height=10}
nbsf <- st_bridges(df = uk_election,
                   geom_col_name = "constituency_name",
                   link_islands_k = 2)
st_quickmap_nb(nbsf,
                    pointsize=0.05, 
                    title = "st_bridges() contiguities",
                    subtitle = "islands linked to k=2 nearest constituencies")

The neighbourhood structure which is created can be either a named list (the default) or a named matrix. Different modelling packages have different requirements for how this information should be presented. Furthermore, we can choose `add_to_dataframe` to be TRUE (the default) to return a dataframe with a column called `nb` which contains the named list or matrix. If FALSE, only the neighbourhood structure itself is returned

These options can be seen in the unexecuted code below:

```{r, eval=FALSE}
nbsf <- st_bridges(df = sf_dataframe,
                  geom_col_name = "the column containing the names of the contiguous areas",
                  remove_islands = T/F,
                  link_islands_k = 1...n,
                  nb_structure = "list"/"matrix",
                  add_to_dataframe = T/F,
                  title = "title",
                  subtitle = "subtitle")

The neighbourhoods are here in a list form:

```{r}

head(nbsf$nb)

But the "nb" column can also be a matrix, and `st_quickmap_nb()` will still return the same map:

```{r, fig.width=12, fig.height=10}
nbsf <- st_bridges(df = uk_election,
                   geom_col_name = "constituency_name",
                   remove_islands = F,
                   link_islands_k = 2,
                   nb_structure = "matrix",
                   add_to_dataframe = T)
st_quickmap_nb(nbsf,
                    pointsize=0.05)

The matrix neighbourhood structure is now of the following form:

```{r}

nbsf$nb[1:10,1:10]

### Editing the contiguities

There are also functions for manually changing the results of a neighbourhood construction. It may be the case that you want to add some additional links or to remove others. For example, you may be aware from local knowledge of connectivities which are not represented by mere contiguity of polygons. The presence of tunnels or bridges across a body of water would be an example of such a situation. The functions `st_manual_join_nb()` and `st_manual_cut_nb()` do this.

#### sfisland::st_check_islands()

To make the use of these functions easier and more intuitive, the function `st_check_islands()` shows us what contiguities have been set up for the islands by `st_bridges()`:

```{r}
nbsf |> st_check_islands()
```

Sometimes, as in this case, certain islands will have more than the specified k neighbours. This is due to the need for symmetry in the structure.

#### st_manual_join_nb() / st_manual_cut_nb()

Let us say we want to change some of these. For example, I will cut the tie between **Isle Of Wight** and **New Forest East** (using their numbers) and also between **Ynys Mon** and **Arfon** (using their names):

```{r}
nbsf |> 
  st_manual_cut_nb(292,378) |>
  st_manual_cut_nb("Ynys Mon","Arfon") |>
  st_check_islands()
```

As extra contiguities can be difficult to distinguish, an extreme case is shown below to demonstrate `st_manual_join_nb()`. **Gower** in South Wales is joined to **St Ives** in Cornwall and then mapped:

```{r}
st_bridges(df = uk_election|> filter(region %in% c("Wales","South West")),
           geom_col_name = "constituency_name",
           link_islands_k = 2)  |> 
  st_manual_join_nb("Gower","St Ives") |> 
  st_quickmap_nb()
```

These manual functions can also, of course, be used to edit any of the previously discussed neighbourhood structures created by `sfdep`.

For example, looking just at Scotland we can use `sfdep::st_nb_lag_cumul()` to get first and second degree neighbours:

```{r}

uk_election |> 
  filter(region == "Scotland") |> 
  mutate(nb = st_contiguity(uk_election$geometry[uk_election$region == "Scotland"]) |> 
           st_nb_lag_cumul(2)) |>  
  st_quickmap_nb()

This does not include the island constituencies. We can chose to include them by first using `st_bridges()` with k=2...

```{r}

uk_election |> 
  filter(region == "Scotland") |> 
  st_bridges(geom_col_name = "constituency_name",
             link_islands_k = 2) |> 
  st_quickmap_nb()

... and then examining the connections which have been made for islands:

```{r}

uk_election |> 
  filter(region == "Scotland") |> 
  st_bridges(geom_col_name = "constituency_name",
             link_islands_k = 2) |> 
  st_check_islands()
```

We can then add these island connections to the output of `sfdep::st_nb_lag_cumul()`:

```{r}

uk_election |> 
  filter(region == "Scotland") |> 
  mutate(nb = st_contiguity(uk_election$geometry[uk_election$region == "Scotland"]) |> 
           st_nb_lag_cumul(2)) |> 
  st_manual_join_nb(47,9) |>
  st_manual_join_nb(47,51) |>
  st_manual_join_nb(47,55) |>
  st_manual_join_nb(51,9) |>
  st_manual_join_nb(51,47) |>
  st_quickmap_nb()

## Modelling & post-functions

Having set up a neighbourhood structure and embedded it as a named list or matrix within the original `sf` dataset as column `nb`, there are some functions to make it easy to quickly perform ICAR smoothing, augment the original dataframe with these predictions, and visualise them.

For example, we can use the `mgcv` package to generate a Markov Random Field ICAR smooth of poor health across the study area. This is done very quickly by using `st_bridges()` to prepare the data, putting that inside the `mgcv` GAM formulation, and then piping into the `st_augment()` and `st_quickmap_preds()` functions.

```{r, fig.width=12, fig.height=10}

prep_data <- st_bridges(uk_election, "constituency_name")

gam(health_not_good ~ s(constituency_name, bs='mrf', xt=list(nb=prep_data$nb), k=100),
        data=prep_data, method="REML") |>
  st_augment(prep_data) |>
  st_quickmap_preds()

An equivalent model, this time smoothing over degree_educated, can be fitted using `brms`. This package requires the neighbourhoods to be in matrix form:

```{r, fig.width=12, fig.height=10}

prep_data2 <- st_bridges(uk_election, "constituency_name", nb_structure = "matrix")

# fit <- brm(degree_educated ~ car(W, gr=constituency_name, type="icar"),
#            data = prep_data2, data2 = list(W=prep_data2$nb),
#            family = gaussian(),file = "brmsfit_degree")

fit <- readRDS("/Users/kevinhoran/Documents/sfislands_prepstuff/brmsfit_degree.rds")

prep_data2$brmsfit <- predict(fit,prep_data2)[,1]

ggplot(prep_data2)+geom_sf(aes(fill=brmsfit), linewidth=0.1) +
  scale_fill_gradient2(low="firebrick4",mid="white",high="darkblue",midpoint = 0) +
  coord_sf(datum=NA) +
  theme_minimal() +
  theme_bw()

More complex models with random effects and multiple smooths are also possible with `mgcv` and the `st_augment()` and `st_quickmap_preds()` functions can handle these and label the columns and maps which are generated appropriately. This is shown below with a model of swing in the 2019 election:

```{r, fig.width=12, fig.height=6}

prep_data3 <- st_bridges(uk_election, "constituency_name") # decide upon the contiguities and add them to the df

model <- gam(con_swing ~ 
               s(region, bs="re") + # region level random intercept
               s(county, bs="re") + # county level random intercept
               s(county, degree_educated, bs="re") + # county level random coefficient
               s(constituency_name, bs='mrf', 
                 xt=list(nb=prep_data3$nb),k=10) + # ICAR constituency ICAR varying coefficients
               s(constituency_name, by=white, bs='mrf', 
                 xt=list(nb=prep_data3$nb),k=10), # ICAR constituency ICAR varying coefficients
             data=prep_data3, method="REML") |> 
  st_augment(prep_data3) |> # pipe into function to get estimates
  st_quickmap_preds() # pipe into this for visualisation

ggarrange(plotlist = model, legend = "none", nrow=1)
```

To see the estimates returned:

```{r}

gam(con_swing ~ 
               s(region, bs="re") + # region level random intercept
               s(county, bs="re") + # county level random intercept
               s(county, degree_educated, bs="re") + # county level random coefficient
               s(constituency_name, bs='mrf', 
                 xt=list(nb=prep_data3$nb),k=10) + # ICAR constituency ICAR varying coefficients
               s(constituency_name, by=white, bs='mrf', 
                 xt=list(nb=prep_data3$nb),k=10), # ICAR constituency ICAR varying coefficients
             data=prep_data3, method="REML") |> 
  st_augment(prep_data3) |> 
  head()

### Back to `guerry`

Returning to the dataset used in `sfdep`, we can easily create a smooth of suicides in France in 1830 as follows. Since there are no islands, we use `st_bridges()` and it will function like `sfdep:st_contiguity()` except that it automatically adds a neighbourhood 'nb' column to the dataframe.

```{r}
prep_data4 <- g |> st_bridges("department")

mod4 <- gam(suicides ~ s(department, bs='mrf', xt=list(nb=prep_data4$nb), k=80),
            data=prep_data4, method="REML") |>
  st_augment(prep_data4) |>
  st_quickmap_preds()
ggarrange(plotlist=mod4)

Or fit a more complex mixed model:

```{r, fig.width=12, fig.height=6}
prep_data5 <- g |> 
  st_bridges("department")

model5 <- gam(donations ~ 
                s(region, bs="re") + 
                s(region, crime_prop, bs="re") + # county level random coefficient
                s(department, bs='mrf', 
                  xt=list(nb=prep_data5$nb),k=20) + # ICAR constituency level varying coefficents
                s(department, by=infants, bs='mrf', 
                  xt=list(nb=prep_data5$nb),k=20), # ICAR constituency level varying coefficents
              data=prep_data5, method="REML") |> 
  st_augment(prep_data5) |> # pipe into function to get estimates
  st_quickmap_preds() # pipe into this for visualisation

ggarrange(plotlist = model5, legend = "none", nrow = 1)
```

>>>>

More details of how to operate these functions are provided in the vignette.
